Comparing probabilistic accounts of probability judgments

### 2021-10-18 paper-rmd/paper-papaja.Rmd
finished reloo

abstract: |
  Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic. However, one mark sometimes held against these theories is that people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These theories fit quite differently into the larger project of Bayesian cognitive science, but their many similarities complicate comparisons of their predictive accuracy. In particular, comparing the models demands a careful accounting of model complexity. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison using information criteria. Compared the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by a modified version of the Bayesian Sampler model that allows for informative priors.

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
header-includes   :
  - \usepackage{tikz}
  - \usetikzlibrary{bayesnet}
---

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. First, Bayesian data analysis allows issues of model complexity to be addressed through comparisons of model fit based on information criteria, such as $WAIC$ and $LOO_{ic}$. Both of these indices approximate leave-one-out cross validation, thereby offering a measure of the ability of a model to capture new unseen data. Importantly, these indices also allow for complexity penalties that directly capture differences in the flexibility of the models, instead of being estimated based solely on the number of distinct parameters [@gelman.etal2014; @vehtari.etal2017]. In addition, the Bayesian framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in a potential reduction in model complexity and a more realistic test of the models.

@@

Zhu, Sanborn, & Chater [-@chater.etal2020] conducted two experiments to compare the PT+N and Bayesian sampler theories. These experiments asked participants to judge the probability of different events in various combinations. Following prior work by Costello and Watts [e.g. -@costello.watts2016; -@costello.watts2018], both experiments focused on the everyday events of different kinds of weather.

Experiment 1 asked about the events [icy, frosty] and [normal, typical] (e.g. "what is the probability that the weather in London is normal and not typical?").  The authors’ goal was to ask about highly correlated events, but the events used are perhaps nearly perfectly correlated. Because the terms used to describe these events are nearly synonymous, there is a concern about the interpretation of the statements evaluated in this experiment. This is especially clear, as the authors note, for disjunctive trials such as "normal or typical," where "or typical" might not be read as a disjunction but rather an elaborative clause.  In light of these concerns, I excluded the disjunctive trials from Experiment 1 from my analyses.

@@

I implement both the Bayesian Sampler and PT+N models as Bayesian models in the probabilistic programming language Numpyro. All code and results are available as supplemental materials (https://github.com/derekpowell/bayesian-sampler).

@@

Recall that Zhu and colleagues (2020) identified a bridging condition relating $\beta$ and $N$ in the Bayesian Sampler model to the $d$ parameter of the PT+N model. To support direct comparisons of the models, I parameterize the Bayesian Sampler model according to the implied $d$  and $d'$, rather than directly according to its $\beta$, $N$, and N' parameters. I constrain $d$ to $[0, 1/3]$ for the Bayesian Sampler model to reflect the assumption that $\beta \in [0, 1]$. This allows the same priors to be used for the corresponding Bayesian Sampler and PT+N models, simplifying their comparison.

@@

I fit each of the models specified above to data from Zhu et al’s [-@zhu.etal2020] Experiment 1 and 2 and computed $LOO_{ic}$ scores for each combination. Compared with BIC, $LOO_{ic}$ offers more sophisticated methods for estimating model complexity, and are generally considered more appropriate in the "$\mathcal{M}$-open" case; situations where we do not know if any of the models being compared are the "true" model [@vehtari.etal2019]. Model posteriors were estimated using the Numpyro [@phan.etal2019] implementation of the No-U-Turn Hamiltonian Markov chain Monte Carlo (MCMC) sampler. For each model, four MCMC chains of 2000 iterations were sampled after 2000 iterations of warmup and all passed convergence tests according to $\hat{R}$ [see @gelman.etal2014a]. Figure 3 below displays the estimated differences in $LOO_{ic}$ scores for each of the models as compared to the best-scoring model.

\begin{figure}[ht]
\centering
\includegraphics[width=6in]{plot_compare.png}
\caption{Model comparison results for data from Experiments 1 and 2. Error bars indicate two standard errors of the estimates. Typically, a difference of greater than two standard errors is taken as clear evidence for the superiority of the lower-scoring model \cite{sivula.etal2020}.}
\end{figure}

Data from Experiment 1 favor "complex" variants of the Bayesian Sampler model compared with the "simple" variants and all versions of the PT+N model. However, there is no clear winner. The beset-scoring model is an unrestricted variant of the Bayesian Sampler allows for people to bring informative priors to the probability judgment task (again, lower scores for $LOO_{ic}$ are better). That is, a Bayesian Sampler model without any restriction on $\beta$, allowing it to range from [0, $\infty$]. However, the complex and complex hierarchical implementations of the Bayesian Sampler assuming uninformative priors (i.e., restricting $\beta \in [0,1]$) have $\text{LOO}_{ic}$ scores within two standard errors of the difference, indicating that these models are also plausible.

Data from Experiment 2 more decisively reveal a single winning model: the hierarchical "unrestricted" implementation of the Bayesian Sampler model allowing for informative priors. Recall, this model may also be seen as a version of the PT+N theory that removes the original theory’s constructive account of conditional probability judgments. Comparing predictions for conditional probability judgments from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from both experiments (see Table 2). These findings suggest that conditioning is better seen as part of the mental model than the probability judgment process.

Figure 4 shows the posterior distributions of the population-level $d$ and $d'$ parameters inferred from the unrestricted Bayesian Sampler model. In Experiment 2, population-level estimates of $d$  and $d'$ are greater than $\frac{1}{3}$, outside the range implied by the assumption of "ignorance priors" in the Bayesian Sampler model. Parameters fit to the data from Experiment 1 are more consistent with this assumption, although a substantial proportion of individual participants’ $d$  and $d'$ estimates also lie outside this range (13 of 59 for d, 26 of 59 for $d'$). The finding that there are clear differences in $d$  and $d'$ estimated across experiments suggest that the mental sampling processes producing estimates vary in the different conditions, either in terms of the number of samples that are drawn, the noise in reading those samples, or the form of the prior distribution assumed by participants in each context.

@@

Compared to the Bayesian Sampler model, the PT+N model with its constructive account of conditional probability judgments has a similar penalty term estimate when fit to Experiment 1, but has a smaller penalty term when fit to Experiment 2, despite having the same parameterization in terms of $d$  and $d'$. Evidently this structural component isn’t a complexifying epicycle, instead it is better seen as a prediction that constrains flexibility. However, it is one that ultimately leads to a worse-fitting model.

@@

```{r table2, echo=FALSE, message=F, warning=F}
t2_df <- read_csv("model-comparison-table.csv")

t2_df %>% 
  select(-`...1`) %>% 
  gather(measure, value, -Experiment, -model) %>% 
  pivot_wider(names_from = c(Experiment, measure), values_from=value,names_sort = T) %>% 
  rename(Model = model) %>%
  kbl(
    caption="Bayesian model comparison results with best scoring model in bold face.", 
    digits=3, 
    booktabs=T,
    col.names = c("Model", "$LOO_{ic}$", "Penalty","$r_{resp}$", "$r_{trial}$", "$LOO_{ic}$", "Penalty","$r_{resp}$", "$r_{trial}$"),
    escape = F
    ) %>%
  add_header_above(c(" ", "Experiment 1"=4, "Experiment 2"=4)) %>% 
  kable_classic(full_width=F) %>% 
  kable_styling(latex_options="scale_down") %>% 
  row_spec(1, bold=TRUE)
```

Finally, it is worth noting that these models provide quite strong overall fits to the data, not just for the query averages, but also for the trial averages across individual participants as seen from the correlations between predicted and observed responses in Table 2.

@@

By a fair margin, the model best accounting for the experimental data from Zhu and colleagues [-@zhu.etal2020] was a version of the Bayesian Sampler model without restriction on the range of its $\beta$ parameters. Alternatively, this model can also be seen as a variant of the PT+N model that removes its account of conditional probability judgments. Thus, what these findings indicate most clearly is that the Bayesian Sampler theory provides a superior account of conditional probability judgments in this experimental task. In keeping with the larger theoretical framework of Bayesian cognitive science, this theory assumes that subjective probabilities underlie people’s probability judgments, and that conditional probability judgments are produced by Bayesian conditioning occurring in their mental models of the events in question, rather than as arising from the probability judgment process [@chater.etal2020; @zhu.etal2020].

Findings from  Zhu and colleagues’ experiments do cast doubt on their proposal that the priors of the Bayesian Sampler model should reflect "ignorance priors," symmetric Beta distributions with $\beta \in [0, 1]$. As a generic prior that would be used across contexts, this class of uninformative priors has an appealing rational basis. Nevertheless, the data suggest that people may bring informative priors to the probability judgment task, indicated by estimated $d$  and $d'$ parameters outside the [1, $1/3$] range implied by "ignorance" priors. It could be that people bring domain-specific priors to judgment tasks, meaning that the most appropriate priors might be dictated by the context in which they make their judgments. This could potentially account for the differences in $d$ and $d'$ observed across Experiments 1 and 2.

@@

Finally, some of the most interesting implications of these models go well beyond the probability judgment task itself: the models both support a probabilistic account of beliefs [@chater.etal2020]. Indeed, by representing the true subjective probabilities as a latent variable, Bayesian data analysis allows those underlying credences to be inferred. Examining the model posteriors here reveals these estimates often come with considerable uncertainty, but at least for some participants they can be estimated with useful levels of precision. Of course, Zhu and colleagues’ [-@zhu.etal2020] experiments were never designed for this purpose. Future research could explore how estimates of people's credences might be made more reliable, and how inferences about these mental probabilities might be integrated with other Bayesian models of reasoning [e.g. @franke.etal2016; @jern.etal2014; @griffiths.tenenbaum2006]. One particularly promising direction could be to integrate these models with formal models of belief revision, which might then shed new light on these fundamental cognitive processes [e.g. @cook.lewandowsky2016; @jern.etal2014; @powell.etal2018].

>>>>

abstract: |
 Bayesian theories of cognitive science hold that cognition is fundamentally probabilistic. However, one mark sometimes held against these theories is that people’s explicit probability judgments often violate the laws of probability. Two recent proposals, the "Probability Theory plus Noise" [@costello.watts2014] and "Bayesian Sampler" [@zhu.etal2020] theories of probability judgments, both seek to account for these biases while maintaining that mental credences are fundamentally probabilistic. These theories fit quite differently into the larger project of Bayesian cognitive science, but their many similarities complicate comparisons of their predictive accuracy. In particular, comparing the models demands a careful accounting of model complexity. Here, I cast these theories into a Bayesian data analysis framework that supports principled model comparison using information criteria. Comparing the fits of both models on data collected by Zhu and colleagues [-@zhu.etal2020] I find the data are best explained by a modified version of the Bayesian Sampler model that allows for informative priors.

@@

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
biblio-style      : "apa"
documentclass     : "apa6"
classoption       : "man"
output:
  papaja::apa6_pdf:
    citation_package: biblatex

header-includes:
  - \usepackage{tikz}
  - \usetikzlibrary{bayesnet}
---

@@

Here, I cast both the Bayesian Sampler and PT+N models into a Bayesian data analysis framework that may permit a more decisive comparison. First, Bayesian data analysis allows issues of model complexity to be addressed through comparisons of model fit based on information criteria, such as ${\text{WAIC}}$ and $\text{LOO}_{ic}$. Both of these indices approximate leave-one-out cross validation, thereby offering a measure of the ability of a model to capture new unseen data. Importantly, these indices also allow for complexity penalties that directly capture differences in the flexibility of the models, instead of being estimated based solely on the number of distinct parameters [@gelman.etal2014; @vehtari.etal2017]. In addition, the Bayesian framework also supports straightforward implementation of hierarchical versions of these models with partial pooling. This allows for information about model parameters to be shared across participants, resulting in a potential reduction in model complexity and a more realistic test of the models.

@@

Zhu, Sanborn, & Chater [-@chater.etal2020] conducted two experiments to compare the PT+N and Bayesian Sampler theories. These experiments asked participants to judge the probability of different events in various combinations. Following prior work by Costello and Watts [e.g. -@costello.watts2016; -@costello.watts2018], both experiments focused on the everyday events of different kinds of weather.

Experiment 1 asked about the events [icy, frosty] and [normal, typical] (e.g. "what is the probability that the weather in London is normal and not typical?").  The authors’ goal was to ask about highly correlated events, but the events used are perhaps nearly perfectly correlated. Because the terms used to describe these events are nearly synonymous, there is a concern about the interpretation of the statements evaluated in this experiment. This is especially clear, as the authors note, for disjunctive query trials such as "normal or typical," where "or typical" might not be read as a disjunction but rather an elaborative clause.  In light of these concerns, I excluded the disjunctive trials from Experiment 1 from my analyses.

@@

I implement several variants of the Bayesian Sampler and PT+N models in a Bayesian framework. These models were implemented in the probabilistic programming language Numpyro. All code and results are available as supplemental materials (https://github.com/derekpowell/bayesian-sampler).

@@

Recall that Zhu and colleagues (2020) identified a bridging condition relating $\beta$ and $N$ in the Bayesian Sampler model to the $d$ parameter of the PT+N model. To support direct comparisons of the models, I parameterize the Bayesian Sampler model according to the implied $d$  and $d'$, rather than directly according to its $\beta$, $N$, and $N'$ parameters. I constrain $d$ to $[0, 1/3]$ for the Bayesian Sampler model to reflect the assumption that $\beta \in [0, 1]$. This allows the same priors to be used for the corresponding Bayesian Sampler and PT+N models, simplifying their comparison.

@@

I fit each of the models specified above to data from Zhu et al’s [-@zhu.etal2020] Experiment 1 and 2 and computed $\text{LOO}_{ic}$ scores for each combination. Compared with BIC, $\text{LOO}_{ic}$ offers a more sophisticated estimation of model complexity and is more appropriate in the "$\mathcal{M}$-open" case; situations where we do not know if any of the models being compared are the "true" model [@vehtari.etal2019]. Model posteriors were estimated using the Numpyro [@phan.etal2019] implementation of the No-U-Turn Hamiltonian Markov chain Monte Carlo (MCMC) sampler. For each model, four MCMC chains of 2000 iterations were sampled after 2000 iterations of warmup and all passed convergence tests according to $\hat{R}$ [see @gelman.etal2014a]. Figure 3 below displays the estimated differences in $\text{LOO}_{ic}$ scores for each of the models as compared to the best-scoring model.

\begin{figure}[ht]
\centering
\includegraphics[width=6in]{plot_compare.png}
\caption[]{Model comparison results for data from Experiments 1 and 2. Error bars indicate two standard errors of the estimates. Typically, a difference of greater than two standard errors is taken as clear evidence for the superiority of the lower-scoring model (\cite{sivula.etal2020}).}
\end{figure}

Data from Experiment 1 favor "complex" variants of the Bayesian Sampler model compared with the "simple" variants and all versions of the PT+N model. However, there is no clear winner. The beset-scoring model is an unrestricted variant of the Bayesian Sampler allows for people to bring informative priors to the probability judgment task (again, lower scores for $\text{LOO}_{ic}$ are better). That is, a Bayesian Sampler model without any restriction on $\beta$, allowing it to range from [0, $\infty$]. However, the complex and complex hierarchical implementations of the Bayesian Sampler assuming uninformative priors (i.e., restricting $\beta \in [0,1]$) have $\text{LOO}_{ic}$ scores within two standard errors of the difference, indicating that these models are also plausible [@sivula.etal2020].

Data from Experiment 2 more decisively reveal a single winning model: the hierarchical "unrestricted" implementation of the Bayesian Sampler model allowing for informative priors.

Figure 4 shows the posterior distributions of the population-level $d$ and $d'$ parameters inferred from the unrestricted Bayesian Sampler model. In Experiment 2, population-level estimates of $d$  and $d'$ are greater than $1/3$, outside the range implied by the assumption of "ignorance priors" in the Bayesian Sampler model. Parameters fit to the data from Experiment 1 are more consistent with this assumption, although a substantial proportion of individual participants’ $d$  and $d'$ estimates also lie outside this range (13 of 59 for d, 26 of 59 for $d'$). The finding that there are clear differences in $d$  and $d'$ estimated across experiments suggest that the mental sampling processes producing estimates vary in the different conditions, either in terms of the number of samples that are drawn, the noise in reading those samples, or the form of the prior distribution assumed by participants in each context.

@@

Recall, the unrestricted Bayesian Sampler model may also be seen as a version of the PT+N theory that removes the original theory’s constructive account of conditional probability judgments. So, comparing the PT+N models to the unrestricted Bayesian Sampler model offers a test of the PT+N's constructive account of conditional probability jugments. Comparing predictions from the unrestricted Bayesian Sampler model and the best-fitting PT+N model, we see that the Bayesian Sampler model better captures these judgments from both experiments (see Table 2). And comparing their prediction specifically for conditional probability judgments, the unrestricted Bayesian Sampler provides a better fit, especially for the modestly-correlated events of experiment 2 (Exp 1: response-level $r$ = .91 vs. .89; query-level $r$ = .98 vs. .97; Exp 2: response-level $r$ = .72 vs. .67; query-level $r$ = .92 vs. .85) These findings suggest that conditioning is better seen as part of the mental model than the probability judgment process. Compared to the Bayesian Sampler model, the PT+N model with its constructive account of conditional probability judgments has a similar penalty term estimate when fit to Experiment 1, but has a smaller penalty term when fit to Experiment 2, despite having the same parameterization in terms of $d$  and $d'$. Evidently this structural component isn’t a complexifying epicycle, instead it is better seen as a prediction that constrains flexibility. However, it is one that ultimately leads to a worse-fitting model.

@@

```{r table2, echo=FALSE, message=F, warning=F}
### NOTE: be sure to rescale elppd_loo to deviance scale to report as loo_ic 10/15/21, 2:37 PM

t2_df <- read_csv("model-comparison-table.csv")

t2_df %>% 
  select(-`...1`) %>% 
  gather(measure, value, -Experiment, -model) %>% 
  pivot_wider(names_from = c(Experiment, measure), values_from=value,names_sort = T) %>% 
  rename(Model = model) %>%
  kbl(
    caption="Bayesian model comparison results with best scoring model in bold face.", 
    digits=3, 
    booktabs=T,
    col.names = c("Model", "$LOO_{ic}$", "Penalty","$r_{resp}$", "$r_{query}$", "$LOO_{ic}$", "Penalty","$r_{resp}$", "$r_{query}$"),
    escape = F
    ) %>%
  add_header_above(c(" ", "Experiment 1"=4, "Experiment 2"=4)) %>% 
  kable_classic(full_width=F) %>% 
  kable_styling(latex_options="scale_down") %>% 
  row_spec(1, bold=TRUE)
```

Finally, it is worth noting that the best of these models provide quite strong overall fits to the data, not just for the query averages, but also for the query averages across individual participants as seen from the correlations between predicted and observed responses in Table 2.

@@

By a fair margin, the model best accounting for the experimental data from Zhu and colleagues [-@zhu.etal2020] was a version of the Bayesian Sampler model without restriction on the range of its $\beta$ parameters. Alternatively, this model can also be seen as a variant of the PT+N model that removes its account of conditional probability judgments. Thus, what these findings indicate most clearly is that the Bayesian Sampler theory provides a superior account of conditional probability judgments in this task. In keeping with the larger theoretical framework of Bayesian cognitive science, the Bayesian Sampler theory assumes that subjective probabilities underlie people’s probability judgments, and that conditional probability judgments are produced by Bayesian conditioning occurring in their mental models of the events in question, rather than as arising from the probability judgment process [@chater.etal2020; @zhu.etal2020].

Zhu and colleagues’ findings do cast some doubt on their proposal that the priors of the Bayesian Sampler model should reflect "ignorance priors," symmetric Beta distributions with $\beta \in [0, 1]$. As a generic prior that would be used across contexts, this class of uninformative priors has an appealing rational basis. Nevertheless, the data suggest that people may bring informative priors to the probability judgment task, indicated by estimated $d$  and $d'$ parameters outside the $[0, 1/3]$ range implied by "ignorance" priors. It could be that people bring domain-specific priors to judgment tasks, meaning that the most appropriate priors might be dictated by the context in which they make their judgments. This could potentially account for the differences in $d$ and $d'$ observed across Experiments 1 and 2.

@@

Finally, some of the most interesting implications of these models go well beyond the probability judgment task itself: the models both support a probabilistic account of beliefs [@chater.etal2020]. Indeed, by representing the "true" subjective probabilities as a latent variable, Bayesian data analysis allows those underlying credences to be inferred. Examining the model posteriors here reveals these estimates often come with considerable uncertainty, but at least for some participants they can be estimated with useful levels of precision. Of course, Zhu and colleagues’ [-@zhu.etal2020] experiments were never designed for this purpose. Future research could explore how estimates of people's credences might be made more reliable, and how inferences about these mental probabilities might be integrated with other Bayesian models of reasoning [e.g. @franke.etal2016; @jern.etal2014; @griffiths.tenenbaum2006]. One particularly promising direction could be to integrate these models with formal models of belief revision, which might then shed new light on these fundamental cognitive processes [e.g. @cook.lewandowsky2016; @jern.etal2014; @powell.etal2018].
